---
title: "R tidyverse dplyr vs. R data.table vs. Python pandas"
description: |
  A short description of the post.
author:
  - name: Hause Lin
    url: {}
date: 11-25-2019
draft: true
categories: 
  - resources
  - data.table
  - python
  - tidyverse
  - dplyr
  - pandas
output:
  radix::radix_article:
    toc: true
    self_contained: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = FALSE, comment = NA, message = FALSE, warning = FALSE)
```


Get source code for this RMarkdown script [here](https://github.com/hauselin/rtutorialsite/blob/master/_posts/2019-07-06-interpreting-interaction-regression-coefficients/interpreting-interaction-regression-coefficients.Rmd).

## Consider being a patron and supporting my work?

[Donate and become a patron](https://donorbox.org/support-my-teaching): If you find value in what I do and have learned something from my site, please consider becoming a patron. It takes me many hours to research, learn, and put together tutorials. Your support really matters.


```{r}
library(dplyr); library(data.table); library(reticulate)
# use_condaenv(condaenv = "python376") # conda python environment
use_condaenv(condaenv = "python376", conda = "/opt/anaconda3/bin/conda")
```

## Loading data

Load data in R

```{r}
iris_df <- as_tibble(iris) # tibble from dplyr
iris_dt <- data.table(iris)  # data.table from data.table
iris_pd <- r_to_py(iris)  # save iris data as an object that can be called/used by python
```

Load data in python

```{python}
import pandas as pd  
iris_pd = r['iris_pd']
```

## View data

**dplyr**

```{r}
iris_df
```

**data.table**

```{r}
iris_dt
```

**pandas**

```{python}
iris_pd
```

## Rename selected columns

```{r}
rename(iris_df, sepal_length = Sepal.Length, sepal_width = Sepal.Width)
setnames(iris_dt, names(iris_dt)[1:2], c("sepal_length", "sepal_width")) # inplace operation
```

```{python}
iris_pd = iris_pd.rename(columns={"Sepal.Length": "sepal_length", "Sepal.Width": "sepal_width"})
```

## Rename all columns

```{r}
iris_df <- rename(iris_df, sepal_length = Sepal.Length, sepal_width = Sepal.Width, petal_length = Petal.Length, petal_width = Petal.Width)

new_names <- tolower(gsub(".", "_", names(iris_dt), fixed = TRUE))
setnames(iris_dt, new_names) # inplace operation
```

```{python}
new_names = iris_pd.columns.str.replace(".", "_").str.lower()
iris_pd.columns = new_names
```

## Select columns using index

```{r}
select(iris_df, 1, 3)
# you can also select and rename at the same time
# select(iris_df, col1 = 1, col3 = 3)
iris_dt[, c(1, 3)]
```

```{python}
iris_pd.iloc[:, [0, 2]]
```

## Slice/select rows using index

```{r}
slice(iris_df, 1, 3)
iris_dt[c(1, 3)]
```

```{python}
iris_pd.iloc[[0, 2]]
```

## Slice/select last row

```{r}
slice(iris_df, n())
iris_dt[.N]
```

```{python}
iris_pd.tail(1)
```

**Score: dplyr: 0, data.table: 0, pandas: 0**

## Select columns using names

```{r}
select(iris_df, sepal_length, petal_length)
iris_dt[, c("sepal_length", "petal_length")]
```

```{python}
iris_pd.loc[:, ["sepal_length", "petal_length"]]
iris_pd[["sepal_length", "petal_length"]]
```

## Select columns with matching names

```{r}
select(iris_df, starts_with("petal"))
iris_dt[, startsWith(names(iris_dt), "petal"), with = FALSE]
```

```{python}
iris_pd.loc[:, iris_pd.columns.str.startswith("petal")]
iris_pd[iris_pd.columns[iris_pd.columns.str.startswith("petal")]] # equivalent to above
```

## Select multiple columns flexibly

```{r}
# dplyr
select(iris_df, ends_with("h"), -1, 5) # ends with h, drop col 1, select col 5

# data.table: not possible in one line!
filt <- endsWith(names(iris_dt), "h")
filt[1] <- FALSE
filt[5] <- TRUE
iris_dt[, filt, with = FALSE]
```

```{python}
# pandas: not possible in one line
filt = iris_pd.columns.str.endswith('h')
filt[0] = False
filt[4] = True
iris_pd.loc[:, filt]
```

**Score: dplyr: 1, data.table: 0, pandas: 0**

## Subset/filter data

```{r}
filter(iris_df, sepal_length < 4.5)
iris_dt[sepal_length < 4.5]
```

```{python}
iris_pd[iris_pd['sepal_length'] < 4.5]
iris_pd.query("sepal_length < 4.5")
```

