---
title: "Analyze multi-subject/trial RT data"
description: Simple tutorial and workflow for analyzing reaction time (RT) and accuracy data from multiple experimental subjects
output:
    radix::radix_article:
        toc: true
        toc_depth: 3
repository_url: https://github.com/hauselin/rtutorialsite/blob/master/0005_analyze_multi_subject_trial_data.Rmd
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
# this chunk is for knitting Rmarkdown; you don't need this chunk if you're not knitting
knitr::opts_chunk$set(echo = TRUE, cache = FALSE, comment = NA, message = FALSE, warning = FALSE, R.options = list(width = 80), tidy = TRUE, tidy.opts = list(width.cutoff = 80))
```

This tutorial standard reaction time and acccuracy data that resemble data collected from standard psychology experiments (though the data have been simulated with the `rwiener` function from the `RWiener` package). This package also assumes you know how to set your working directory; if not, check out my intro tutorials [here](https://hausetutorials.netlify.com/0001_rbasics.html#workingcurrent-directory-where-are-you-and-where-should-you-be-getwd). 
## Getting the data and code

* Download Rmarkdown script for this tutorial [here](https://raw.githubusercontent.com/hauselin/rtutorialsite/master/0005_analyze_multi_subject_trial_data.Rmd).
* Download data for all my tutorials [here](https://www.dropbox.com/sh/2ix85ij7chfuqdg/AABpkIMymvjOE5Dh9iV7SULga?dl=1).
* Inside the data folder, you should see a `rt_acc_data_raw` folder that contains three subjects' data, which we will analyze in this tutorial.
* Leave the folders and files in the unzipped folder as they are because we'll use R to read the csv files into our R environment.

## Dataset information

* 3 subjects: 30, 35, 39
* Completed a task where reaction time (rt) and accuracy (acc) were collected.
* 2 blocks in the experiment (1, 2)
* 40 trials per block; 4 blocks in total; 80 trials in total
* 2 experimental manipulations: reward available on trial (high vs low); instructions for block (focus on responding carefully [caution] vs focusing on responding quickly (speed))

### Variables 

* subject: subject id
* completed: whether subject completed study (1, 0)
* rt: reaction time
* acc: accuracy
* reward: reward available on trial (high or low)
* instructions: block instructions (caution or speed)
* block: block number (1, 2)
* trialinblock: trial in block (1 to 4 0)
* overalltrial: overall trial number (1 to 80)

## Organizing folders/directory

This is how my directories/folders/files look like. You should make sure yours look similar before you continue.

— 0005_analyze_multi_subject_trial_data.Rmd

——— data

————— rt_acc_data_raw

——————— subject_30.csv

——————— subject_35.csv

——————— subject_39.csv

## Load libraries

```{r}
library(tidyverse); library(data.table); library(dtplyr); library(lme4); library(lmerTest); library(ggbeeswarm); library(hausekeep)
```

<aside>
`hausekeep` is my package. It contains a few helper functions. You may or may not want to use them. If you want to, install the package by running `devtools::install_github("hauselin/hausekeep")` (you might need to install `devtools` package first). 
</aside>

## Read all subjects' data into R

```{r}
(files <- list.files(path = "./data/rt_acc_data_raw", pattern = "subject_", full.names = TRUE)) # find matching file names
dt1 <- lapply(files, fread) %>% bind_rows() %>% tbl_dt() # read data, combine items in list, convert to data.table and tibble
```

Note and explanation: `lapply` loops through each item in `files` and applies the `fread` function to each item. `bind_rows` combines all the dataframes stored as separate items in the list, and `tbl_dt()` converts the merged dataframe into `data.table` and `tibble` class.

## Summarize experimental design

```{r}
dt1[, .N, by = subject] # rows (trials) per subject
dt1[, .N, keyby = .(subject, block)] # blocks per subject and rows/trials per block
dt1[, .N, keyby = .(subject, instructions, reward)] # row (trials) per subject for instructions/reward conditions
```

## Remove outlier trials

```{r}
dt1[rt < 0.2, `:=` (rt = NA, acc = NA)] # if rt < 0.2, convert rt and acc to NA (too fast response on that trial)
dt1[rt > 3.0, `:=` (rt = NA, acc = NA)] # if rt > 3.0, convert rt and acc to NA (too slow response on that trial)
summary(dt1$rt)
summary(dt1$acc)
```

## Compute mean RT and ACC for each subject, for each condition

```{r}
dt1_avg <- dt1[, .(rt = mean(rt, na.rm = T), acc = mean(acc, na.rm = T)), keyby = .(subject, instructions, reward)]
dt1_avg
```

## Compute grand average RT and ACC (and within-subject error bars)

```{r}
dt1_grdavg <- seWithin(data = dt1_avg, measurevar = c("rt", "acc"), withinvars = c("instructions", "reward"), idvar = "subject")
```

<aside>
`seWithin` computes within-subject error bars and is a function from my `hausekeep` package.
</aside>

## Plot reaction time 

```{r}
plot_rt <- ggplot(dt1_grdavg$rt, aes(instructions, rt, col = reward)) +
  geom_quasirandom(data = dt1_avg, alpha = 0.2, dodge = 1, size = 0.6) +
  geom_point(position = position_dodge(1), shape = 95, size = 6) +
  geom_errorbar(aes(ymin = rt - ci, ymax = rt + ci), position = position_dodge(1), width = 0, size = 1.1) +
  labs(y = "reaction time (95% CI)")
plot_rt
```

## Plot accuracy

```{r}
plot_acc <- ggplot(dt1_grdavg$acc, aes(instructions, acc, col = reward)) +
  geom_quasirandom(data = dt1_avg, alpha = 0.2, dodge = 1, size = 0.6) +
  geom_point(position = position_dodge(1), shape = 95, size = 6) +
  geom_errorbar(aes(ymin = acc - ci, ymax = acc + ci), position = position_dodge(1), width = 0, size = 1.1) +
  scale_y_continuous(limits = c(0, 1.2), breaks = seq(0, 1.0, 0.25)) +
  labs(y = "accuracy (95% CI)")
plot_acc
```

## Fit linear mixed models to single-trial RT and ACC data

```{r}
model_rt <- lmer(rt ~ instructions + reward + (1 | subject), data = dt1)
summary(model_rt)
summaryh(model_rt) # summary function from my hausekeep package

model_acc <- glmer(acc ~ instructions + reward + (1 | subject), data = dt1, family = binomial)
summary(model_acc)
summaryh(model_acc) # summary function from my hausekeep package
```
