---
title: "Use median absolute deviation instead of z-score to detect outliers"
description: |
  Why is it bad to use z-scores to detect outliers and why you should use median absolute deviation instead
preview: https://raw.githubusercontent.com/hauselin/rtutorialsite/master/attachments/mad_zscore_outliers.png
author:
  - name: Hause Lin
    url: {}
date: 09-07-2019
repository_url: https://github.com/hauselin/rtutorialsite/blob/master/_posts/2019-10-07-outlier-detection-with-median-absolute-deviation/outlier-detection-with-median-absolute-deviation.Rmd
categories: 
  - outlier detection
  - hausekeep
output:
  radix::radix_article:
    toc: true
    self_contained: false
editor_options: 
  chunk_output_type: console
bibliography: biblio.bib
---

```{r setup, include=FALSE}
library(roxygen2)
knitr::opts_chunk$set(echo = TRUE, cache = FALSE, comment = NA, message = FALSE, warning = FALSE)
```

Get source code for this RMarkdown script [here](https://github.com/hauselin/rtutorialsite/blob/master/_posts/2019-10-07-outlier-detection-with-median-absolute-deviation/outlier-detection-with-median-absolute-deviation.Rmd).

## Consider being a patron and supporting my work?

[Donate and become a patron](https://donorbox.org/support-my-teaching): If you find value in what I do and have learned something from my site, please consider becoming a patron. It takes me many hours to research, learn, and put together tutorials. Your support really matters.

## Problems with z-score approach

Consider the small set of numbers assigned to variable `scores` below. Just by eyeballing the data, it looks like there's one outlier: 1000

```{r}
scores <- c(-3, 1, 3, 3, 6, 8, 10, 10, 1000)
mean(scores) # mean 
sd(scores) # sd
boxplot(scores) # 1000 looks like an extreme outlier
```

What are the standard or z-scores for each value? 

$$z_{i} = \frac{x_{i}-\overline{x}}{\sigma}$$
where $z_{i}$ is the z-score for a particular score, $x_{i}$ is a particular score, $\overline{x}$ is the mean of all scores, and $\sigma$ is the standard deviation of all scores.

The equation above expressed in code:

```{r}
scores_z <- (scores - mean(scores)) / sd(scores)
scores_z  # 1000 has a z-score < 3
```

The z-score of the largest value is `r tail(scores_z, 1)`, which is relatively small, considering researchers tend to consider scores as outliers only if they have z-scores 3 or larger. 

### Outliers bias the mean and standard deviation

So we seem to have a problem here: By eyeballing the scores, we intuitively know that 1000 should be an outlier, but the z-score outlier detection approach suggests 1000 isn't an outlierand we shouldn't remove it Of course, you could set your exclusion criterion to "scores with z-scores 2.5 (rather than 3.0) or greater will be considered outliers". But changing the criterion arbitrarily doesn't address the main problem:

* extremely negative/positive scores bias the mean and standard deviation, affecting the resulting z-score

```{r}
mean(scores) # original mean
mean(scores[1:8]) # mean after excluding extreme value

sd(scores) # original sd
sd(scores[1:8]) # sd after excluding extreme value
```

### Function from hausekeep package to compute z-scores

To make it easier to compute z-scores, detect outliers, and remove extreme values (based on your cut-off, 1.96, 2.5 or 3 or whatever), I've created the function `outliersZ()`, which is available in my `hausekeep` package. When you run the function, it tells you how many outliers were detected and what they've been replaced by (default replaces them with `NA`). 

```{r}
library(hausekeep)
outliersZ(scores, zCutOff = 3.0) # replace values with z-scores greater than ± 3
outliersZ(scores, showZValues = TRUE) # show z values (default = FALSE)
outliersZ(scores, zCutOff = 0.35)  # note that default zCutOff is 1.96
outliersZ(scores, replaceOutliersWith = -9999) # replace outlier values with -9999
```

<aside>
You might notice the z-scores returned by `outliersZ()` are different from what we computed above. That's because the `sd()` function in R by default uses `n-1` in the denominator when calculating standard deviation, whereas my function uses `n`.
</aside>

For more information, type `?outliersZ` in your console. If you want to see how the `outliersZ()` function is defined, just type `outliersZ` in your console and you'll see the source code.

## Robust solution: median absolute deviation

Instead of using z-scores to detect outliers (which is problematic for various reasons shown above), we can instead use a simple and robust alternative that isn't influenced by extreme outlier values: median absolute deviaion [@Leys2019;@Leys2013]. See [Wikipedia article on median absolute deviation](https://en.wikipedia.org/wiki/Median_absolute_deviation).

Before we can compute the median absolute deviation, we need the median: 

```{r}
scores_median <- median(scores)
scores_median
```

Next we subtract the median from each value and get the absolute values:

```{r}
scores_median_absolute_deviation <- abs(scores - scores_median)
scores_median_absolute_deviation
```

Next we get the median of the median absolute deviations:

```{r}
median_median_absolute_deviations <- median(scores_median_absolute_deviation)
median_median_absolute_deviations
```

Next, we can compute the median of the absolute deviations (MAD):

```{r}
scores_mad <- median_median_absolute_deviations * 1.4826
scores_mad
```

This value is conceptually equivalent to standard deviation, but is used when computing median.

And why 1.4826? It's a constant linked to the assumption of normality of the data, disregarding the abnormality induced by outliers [@Rousseeuw1993].

You can actually easily compute the median of the absolute deviations by calling the `mad()` function. I computed it step-by-step above to show you how it's done. Typically, you can simply call the `mad()` function and provide your raw values as input.

```{r}
mad(scores)  # save as our manually computed scores_mad 
```

Finally, you can compute how much each value deviated:

```{r}
scores_deviation <- (scores - scores_median) / scores_mad
scores_deviation # note that the value of 1000 has a huge deviation of 167!
```

Again, the idea is conceptually similar to computed z-scores: for each value, subtract the median from it, and divide by the median of the absolute deviations.

Using this robust approach, the largest value 1000 in our set of values (`r scores`) has a huge deviation of `r scores_deviation[length(scores_deviation)]`. Regardless of what criteria you use (2.0, 2.5, 3.0 are all common cut-offs you could use), 1000 is so large that it'll have to be excluded (consistent with our intuition).


### Function from hausekeep package to compute median absolute deviation

To help detect and remove outliers using this robust approach, I've created the function `outliersMAD()`, which is available in my `hausekeep` package. When you run the function, it tells you how many outliers were detected and what they've been replaced by (default replaces them with `NA`). 

```{r}
library(hausekeep)
outliersMAD(scores, MADCutOff = 3.0) # replace values with deviations greater than ± 3
outliersMAD(scores, showMADValues = TRUE) # show deviation (default = FALSE)
outliersMAD(scores, MADCutOff = 0.6)  # note that default cut-off value is 3.0
outliersMAD(scores, replaceOutliersWith = -9999) # replace outlier values with -9999
```

For more information, type `?outliersMAD` in your console. If you want to see how the `outliersMAD()` function is defined, just type `outliersMAD` in your console and you'll see the source code.

Here's how you would use the function typically:

```{r}
scores_outliers_removed <- outliersMAD(scores)
scores_outliers_removed
```

```{r}
boxplot(scores_outliers_removed)
```

## One final example

```{r}
scores <- c(-5, -2, 4, 8, 55, 100)
```

Use z-score method with 1.96 as cut-off:

```{r}
scores_removeoutliers_zscore <- outliersZ(scores) # uses 1.96 as default cutoff
scores_removeoutliers_zscore
```

Use median absolute deviation method with 3.00 as cut-off:

```{r}
scores_removeoutliers_mad <- outliersMAD(scores) # uses 3.00 as default cutoff
scores_removeoutliers_mad
```

```{r}
par(mfrow = c(1, 3))  # set figure to plot 1 row, 3 columns
boxplot(scores, main = "raw values")
boxplot(scores_removeoutliers_zscore, main = "after z-score outlier removal")
boxplot(scores_removeoutliers_mad, main = "after mad outlier removal")
```

## Support my work

[Support my work and become a patron here](https://donorbox.org/support-my-teaching)!
