---
title: "Parallel programming"
description: "test"
preview: https://raw.githubusercontent.com/hauselin/rtutorialsite/master/attachments/plot_dotdist.gif
author:
  - name: Hause Lin
date: 10-06s-2019
output:
  radix::radix_article:
    toc: true
    self_contained: false
draft: true
categories: 
  - foreach
  - parallel
repository_url: https://github.com/hauselin/rtutorialsite/blob/master/_posts/2019-02-22-why-we-should-never-use-barplots-use-geomquasirandom-instead/why-we-should-never-use-barplots-use-geomquasirandom-instead.Rmd
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(roxygen2)
knitr::opts_chunk$set(echo = TRUE)
```

Get source code for this RMarkdown script [here](https://github.com/hauselin/rtutorialsite/blob/master/_posts/2019-02-22-why-we-should-never-use-barplots-use-geomquasirandom-instead/why-we-should-never-use-barplots-use-geomquasirandom-instead.Rmd).

## Consider being a patron and supporting my work?

[Donate and become a patron](https://donorbox.org/support-my-teaching): If you find value in what I do and have learned something from my site, please consider becoming a patron. It takes me many hours to research, learn, and put together tutorials. Your support really matters.

## Load `foreach` library

```{r}
library(foreach)
```









## Sequential.serial processing (non-parallel)

### Basic

```{r}
foreach(nvalues = c(1, 2, 3)) %do% rnorm(n = nvalues)
```



```{r}
foreach(nvalues = c(4, 8), meanvalues = c(10, 500), sdvalues = c(1, 500)) %do% 
  rnorm(n = nvalues, mean = meanvalues, sd = sdvalues)
```


### Combining results

```{r}
foreach(nvalues = c(4, 4), meanvalues = c(10, 500), sdvalues = c(1, 500), .combine = rbind) %do% 
  rnorm(n = nvalues, mean = meanvalues, sd = sdvalues)
```


```{r}
foreach(nvalues = c(4, 4), meanvalues = c(10, 500), sdvalues = c(1, 500), .combine = "+") %do% 
  rnorm(n = nvalues, mean = meanvalues, sd = sdvalues)
```

### List comprehension

```{r}
foreach(x = sample(1:1000, 10), .combine = c) %:%
  when(x %% 3 == 0 || x %% 5 == 0) %do%
  x
```


## Parallel processing

`doParallel`, `doFuture`, or `doSEQ`backends

```{r}
library(doParallel)
clust <- makeCluster(3)
registerDoParallel(clust)
foreach(nvalues = c(1, 2, 3)) %dopar% rnorm(n = nvalues)
```

### doFuture 

```{r}
library(doFuture)
registerDoFuture()
plan(cluster, workers = 2)
foreach(nvalues = c(1, 2, 3), .combine = c) %dopar% rnorm(n = nvalues)
```


multicore plan

```{r}
plan(multicore)
foreach(nvalues = c(1, 2, 3), .combine = c) %dopar% rnorm(n = nvalues)
```



## Resources

https://privefl.github.io/blog/a-guide-to-parallelism-in-r/
https://nceas.github.io/oss-lessons/parallel-computing-in-r/parallel-computing-in-r.html
https://campus.datacamp.com/courses/parallel-programming-in-r/foreach-futureapply-and-load-balancing?ex=6
https://cran.r-project.org/web/packages/foreach/vignettes/nested.pdf


## Support my work

[Support my work and become a patron here](https://donorbox.org/support-my-teaching)!
