---
title: "Running R and Python with reticulate (RStudio preview version 1.2)"
description: |
  Running R and Python seamlessly in RStudio 1.2 with reticulate package.
author:
  - name: Hause Lin
date: 02-22-2019
output:
  radix::radix_article:
    self_contained: false
draft: true
categories: 
  - reticulate
  - python
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(roxygen2)
knitr::opts_chunk$set(echo = TRUE)
```

Get source code for this RMarkdown script [here](https://github.com/hauselin/rtutorialsite/blob/master/_posts/2019-02-22-why-we-should-never-use-barplots-use-geomquasirandom-instead/why-we-should-never-use-barplots-use-geomquasirandom-instead.Rmd).

The next major version of RStudio (1.2) will provide much more support for Python in RStudio. I tried running Python and R code chunks in an RMarkdown document and everything ran really seamlessly!

```{r loading packages, results="hide", message=FALSE, warning=FALSE}
# load packages
library(data.table); library(tidyverse); library(gganimate); 
library(reticulate) # runs python
```

<aside>r chunk</aside>

The [reticulate](https://rstudio.github.io/reticulate/) package lets us run Python and R in RStudio. 

```{r}
Sys.which("python") # default version of Python found on your PATH
```

<aside>r chunk</aside>

If you have Python conda environments, you can activate that environment in RStudio. Here, I'm activating my conda environment `python368`.

```{r}
use_condaenv("python368") # use my conda environment 
```

<aside>r chunk</aside>

Let's import Python packages `pandas`!

```{python}
import pandas as pd
```

<aside>python chunk</aside>

Create a pandas dataframe in Python.

```{python}
data = {'col_1': [3, 2, 1, 0], 'col_2': ['a', 'b', 'c', 'd']}
df1 = pd.DataFrame(data)
print(df1)
```

<aside>python chunk</aside>

Create the same dataframe in R

```{r}
df2 <- tibble(col_1 = c(3, 2, 1, 0), col_2 = c("a", "b", "c", "d"))
print(df2)
```

<aside>r chunk</aside>

Loop in Python (note Python starts indexing at 0)

```{python}
for i in range(3):
  print(i)
```

<aside>python chunk</aside>

Loop in R (R starts indexing at 1)

```{r}
for (i in 1:3) {
  print(i)
}
```

<aside>r chunk</aside>

Plot the iris dataset in Python using `seaborn` and `matplotlib.pyplot`

```{python}
import matplotlib.pyplot as plt
import seaborn as sns
iris = sns.load_dataset("iris")
iris["ID"] = iris.index
iris["ratio"] = iris["sepal_length"]/iris["sepal_width"]
sns.lmplot(x="ID", y="ratio", data=iris, hue="species", fit_reg=False, legend=False)
plt.legend()
plt.show()
```

<aside>python chunk</aside>

Plot the iris data in R using `ggplot2`

```{r}
iris$index <- 1:nrow(iris)
ggplot(iris, aes(index, Sepal.Length/Sepal.Width, col = Species)) + 
  geom_point() +
  labs(x = "ID", y = "ratio", col = "species")
```

<aside>r chunk</aside>
