---
title: "Grand-mean and group-mean centering in mixed models"
description: |
  Understand the effects of grand-mean and group-mean centering in mixed models
preview: ./dag0.png
author:
  - name: Hause Lin
date: 03-01-2020
draft: true
categories: 
  - regression
  - mixed models
  - general linear model
bibliography: biblio.bib
output:
  radix::radix_article:
    toc: true
    self_contained: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = FALSE, comment = NA, message = FALSE, warning = FALSE)
```

Get source code for this RMarkdown script [here](https://raw.githubusercontent.com/hauselin/rtutorialsite/master/_posts/2020-04-11-causal-inference-and-lords-paradox-change-score-or-covariate/causal-inference-and-lords-paradox-change-score-or-covariate.Rmd).

## Consider being a patron and supporting my work?

[Donate and become a patron](https://donorbox.org/support-my-teaching): If you find value in what I do and have learned something from my site, please consider becoming a patron. It takes me many hours to research, learn, and put together tutorials. Your support really matters.


```{r}
# load libraries for analyses
library(data.table); library(performance); library(multilevel); library(lme4)
```

```{r}

url <- "https://raw.githubusercontent.com/hauselin/rtutorialsite/master/data/simpsonsParadox.csv"
dt1 <- fread(url)
```

```{r centering}
dt1[, iq_grandc := iq - mean(iq, na.rm = T)]

dt1[, iq_groupc := iq - mean(iq, na.rm = T), by = class]

dt1[, iq_group_mean := mean(iq, na.rm = T), by = class]

dt1[, iq_group_meanc := iq_group_mean - mean(iq_group_mean, na.rm = T)]
```


```{r}
m0 <- lmer(grades ~ iq + (1 | class), dt1)
m1 <- lmer(grades ~ iq_grandc + (1 | class), dt1)
m2 <- lmer(grades ~ iq_groupc + (1 | class), dt1)
m3 <- lmer(grades ~ iq_groupc + iq_group_meanc + (1 | class), dt1)

compare_performance(m0, m1, m2, m3)

library(sjPlot)
tab_model(m0, m1)
```







## Resources

If you want to learn more, read Pearl's[@Pearl] paper and check out the resources below by [Michael Clark](https://m-clark.github.io/).

* [How to center in multilevel models](http://philippmasur.de/blog/2018/05/23/how-to-center-in-multilevel-models/)

## Support my work

[Support my work and become a patron here](https://donorbox.org/support-my-teaching)!
